---
title: "Time Series Analysis"
author: "Aleksandra Kalisz"
date: "2023-05-08"
output: html_document
---

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
library(ggfortify)
library(forecast)
library(stats)
library(smooth)
library(caret)
library(e1071)
library(zoo)
library(fpp2)
library(forecast)
library(tseries)

```



```{r}
# Reading the data

month_temp.df <- read.csv("nitm18442004.csv", fileEncoding = "UTF-8-BOM")
year_temp.df <- read.csv("nity18442004.csv", fileEncoding = "UTF-8-BOM")
tail(month_temp.df)
tail(year_temp.df)


#Summarizing the data
glimpse(month_temp.df)
summary(month_temp.df)

glimpse(year_temp.df)
summary(year_temp.df)
```

```{r}
# Rename column from x to Temperature_y
colnames(year_temp.df)[which(names(year_temp.df)=="x")] <- "Temperature_y"
# Print the updated data frame
print(year_temp.df)


# Rename column x to Temperature_m
colnames(month_temp.df)[which(names(month_temp.df)=="x")] <- "Temperature_m"
# Print the updated data frame
print(month_temp.df)

# Check for missing values
print(colSums(is.na(year_temp.df)))
print(colSums(is.na(month_temp.df)))




```
```{r}
# placing data into time series using ts() function

month_ts <- ts(month_temp.df, start=1844, frequency=12)
year_ts<- ts(year_temp.df,  start=1844,end = 2004)

print(month_ts)
print(year_ts)
```

```{r}

# Performing ADF test
adf.test(month_ts)
adf.test(year_ts)
```

```{r}
# plotting timeseries

plot(month_ts, type = "l", main = "Month Time Series")
plot(year_ts, type = "l", main = "Year Time Series")
```
```{r}
# Creating Seasonal plot of the month time series: This can be used to visualize any seasonal patterns in the data.

seasonplot(month_ts, main = "Seasonal Plot of Month Time Series")

```

```{r}
#Seasonal plot #2
ggseasonplot(month_ts, year.labels=TRUE, year.labels.left=TRUE)+
  ylab("Temperature_m")+
  ggtitle("Seasonal Plot")
 
     
  

```
```{r}
#Seasonal distribution plot #3
ggsubseriesplot(month_ts)+
  ylab("Temperature_m")+
  ggtitle("Seasonal distribution")
```
Smoothing time series- moving averages
```{r}
library(fpp2)
#Plotting Simple moving average for Year


plot(year_ts, main="Raw Time Series Year")
ma(year_ts,6)
ma(year_ts,12)

autoplot(year_ts)+
  autolayer(ma(year_ts,6))+
   autolayer(ma(year_ts,12))
  ggtsdisplay(year_ts)
  



#Plotting Simple moving average for Month


plot(month_ts, main="Raw Time Series Month")
month_smooth<-ma(month_ts,36)
plot(month_smooth)


```
By sliding the window across the data, a new set of smoothed values is obtained, which reduces the impact of random fluctuations and highlights the underlying trends in the data. The choice of window width is important, as it can affect the degree of smoothing achieved and the level of detail preserved in the data.


Seasonal Decomposition using decompose() of multicaptive
```{r}

fit.decmult <- decompose(month_ts, type = "multiplicative")
fit.decmult
plot(fit.decmult)


```

Seasonal Decomposition using decompose() of additive
```{r}

# Multiplicative model for seasonal decomposition Additive
fit.decadd <- decompose(month_ts, type = "additive")
fit.decadd
plot(fit.decadd)

```


Mean Model, Naive Model, SNaive
```{r}

#Mean Model Month
fcast.mean.m<- meanf(month_ts, h=4)
summary(fcast.mean.m)
plot(fcast.mean.m)

# Mean Model for year

fcast.mean.y<- meanf(year_ts, h=4)
summary(fcast.mean.y)
plot(fcast.mean.y)
#############################################
# Naive Model for month
fcast.naive.m<-naive(month_ts, h=3)
summary(fcast.naive.m)
plot(fcast.naive.m)
# where h= 3 periods ahead and look at RMSE value , which is the lowest

# Naive model for Year
fcast.naive.y<-naive(year_ts, h=3)
summary(fcast.naive.y)
plot(fcast.naive.y)

################################################

# seasonal Naive Model Month
fcast.seasonalnaive.m<-snaive(month_ts, h=3)
summary(fcast.seasonalnaive.m)
plot(fcast.seasonalnaive.m)

# seasonal Naive Model year
fcast.seasonalnaive.y<-snaive(year_ts, h=3)
summary(fcast.seasonalnaive.y)
plot(fcast.seasonalnaive.y)
```

Holt's Model
```{r}

 # forecast using a Holt's method
holt <- holt(year_ts, h = 12)
summary(holt)
# Plot the results
plot(holt)

# ETS
holt2 <- window(year_ts, start = 1844)
holt2<- ets(holt2, model = "AAN")
holt2
forecast(holt,5)




```


Holt - Winters Method

```{r}
# Holt Winters Method
holt_winters <- hw(month_ts, seasonal="additive")
autoplot(holt_winters)
summary(holt_winters)


```


CHECKING RESIDUALS
```{r}
checkresiduals(month_ts)
```

```{r}
checkresiduals(year_ts)
```

AUTOCORRELATION AND PARTIAL AUTOCORRELATION
```{r}
Acf(month_ts)
Pacf(month_ts)
Acf(year_ts)
Pacf(year_ts)
```


ARIMA
```{r}
# Arima on month_ts
arima_month <- arima(month_ts, c(1,0,1))
arima_month
summary(arima_month)

# Arima on year_ts
arima_year <- arima(year_ts, c(1,0,1))
arima_month
summary(arima_year)
```

```{r}
# Perform the Ljung-Box test on the residuals of the model
ljung_box <- Box.test(arima_month$residuals, type = "Ljung-Box")

# Print the test results
print(ljung_box)

# Perform the Ljung-Box test on the residuals of the model
ljung_box1 <- Box.test(arima_year$residuals, type = "Ljung-Box")

# Print the test results
print(ljung_box1)
```
```{r}
# forecasting using Arima
forecast_month <- forecast(arima_month,12)
forecast_year <- forecast(arima_year,1)
summary(forecast_month)
summary(forecast_year)

plot(forecast_month)
plot(forecast_year)
```
AUTOMATIC ARIMA

```{r}
#Applying auto Arima to compare
auto_arima <- auto.arima(month_ts, seasonal=FALSE)
auto_arima
summary(auto_arima)


auto_arima1 <- auto.arima(year_ts)
auto_arima1
summary(auto_arima1)

```

```{r}
#Forecasting using auto Arima
forecast (auto_arima)
forecast(auto_arima1)

```
```{r}
#Box text for Auto Arima
Box.test(auto_arima$residuals,type = "Ljung-Box")
Box.test(auto_arima1$residuals,type = "Ljung-Box")

```

SARIMA

```{r}

# Fit a Seasonal ARIMA model
sarima_model <- arima(year_ts, order = c(1, 1, 1), seasonal = list(order = c(0, 1, 1), period = 12))
# View the model summary
summary(sarima_model) # ( NOT SEASONAL)



sarima_model1 <- arima(month_ts, order = c(1, 1, 1), seasonal = list(order = c(0, 1, 1), period = 12))
# View the model summary
summary(sarima_model1)

```

SPLITTING DATA

```{r}
# split monthly temperature data into training and test sets
train_month_ts <- window(month_ts, end=c(2003,12))
test_month_ts <- window(month_ts, start=c(2004,1), end=c(2004,12))

# split yearly temperature data into training and test sets
train_year_ts <- window(year_ts, end=2003)
test_year_ts <- window(year_ts, start=2004, end=2004)

```

To forecast the average temperatures for 2004 using the monthly temperature data, we can fit an appropriate ARIMA model to the training set and use it to make forecasts for the next 12 months. 
```{r}
# Fit a SARIMA model to monthly training data
sarima_month_train <- arima(train_month_ts, order=c(2,0,1), seasonal=c(1,1,0))

# Forecast next 12 months using the SARIMA model
forecast_month <- forecast(sarima_month_train, h=12)

# Plot the forecast and actual data for 2004
plot(forecast_month, main="Monthly Temperature Forecast for 2004 SARIMA")
lines(test_month_ts, col="red")
legend("topleft", legend=c("Monthly Forecast", "Test"), col=c("blue", "red"), lty=1)


```
Holt_winters Forecast
```{r}
# Fit a Holt-Winters model to monthly training data
hw_month_train <- HoltWinters(train_month_ts)

# forecast next 12 months using the Holt-Winters model
forecast_month <- forecast(hw_month_train, h=12)

# plot the forecast and actual data for 2004
plot(forecast_month, main="Monthly Temperature Forecast for 2004 Holt WInters")
lines(test_month_ts, col="red")
legend("topleft", legend=c("Monthly Forecast", "Test"), col=c("blue", "red"), lty=1)

```


```{r}
# fit a seasonal ARIMA model to yearly training data
arima_year_train <- auto.arima(train_year_ts)

# forecast next year using the SARIMA model
forecast_year <- forecast(arima_year_train, h=1)

# plot the forecast and actual data for 2004
plot(forecast_year, main="Yearly Temperature Forecast for 2004")
points(test_year_ts, col="red", pch=20)
legend("bottomleft", legend=c( "Test"), col=c("red"), lty=1, pch=c(1,20))

```

```{r}
# subset monthly temperature data for year 2004
month_2004 <- window(month_ts, start=c(2004,1), end=c(2004,12))
forecast_month <- forecast(sarima_month_train, h=12)

# plot monthly temperature data for year 2004
plot(month_2004, main="Monthly Temperature Data for 2004")


```


```{r}
year_ts <- ts(year_temp.df, start=1844, end=2004)
temp_range <- year_ts[149:161]

# plot the temperature range
plot(temp_range, type="b", main="Yearly Actual Temperature Range for 2004")

# add lines connecting the points
lines(temp_range, col="blue")

# add labels for x and y axes
xlab("Year")
ylab("Temperature")


```




```{r}
# plot the forecast and actual data for 2004
plot(forecast_month, main="Monthly Temperature Forecast for 2004")
lines(month_ts, col="blue")
lines(test_month_ts, col="red")
legend("topleft", legend=c("Forecast", "Original", "Test"), col=c("green", "blue", "red"), lty=1)

# fit a SARIMA model to yearly training data
sarima_year_train <- auto.arima(train_year_ts, seasonal=TRUE)

```



```{r}

# fit an auto.arima model to yearly temperature data
arima_year <- auto.arima(year_ts)

# forecast next year using the auto.arima model
forecast_year <- forecast(arima_year, h=length(test_year_ts))

# plot the forecast and actual data for 2004
plot(forecast_year, main="Yearly Temperature Forecast for 2004")
points(year_ts, col="blue", pch=20)
points(test_year_ts, col="red", pch=20)
lines(forecast_year$mean, col="green")
legend("bottomleft", legend=c("Forecast", "Original", "Test"), col=c("green", "blue", "red"), lty=1, pch=c(1,20,20))

```






